Designing modular software: a case study in introductory statistics

### 2016-09-09 jcgs-paper-2016/paper.Rmd
Restructure the section headers

Modern web technologies have enabled a new generation of software packages that reside solely on the web, which eliminates the issue of local installation. Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. Several recent software packages have built up Shiny to provide a web-based system based on R. One such package is iNZight Lite [@inzight] which attempts to expose students to data analysis without requiring programming knowledge. Like most web-based systems, this does not include reproducible R code, which limits its useful in a scientific or academic setting. Another package is called Radiant [@radiant]

XXX Say something about reproducibility here - it has to be a goal to make an analysis reproducibility (baggerly cancer trials, forensic bioinformatics)

In our view, a modern software system should be **modular**, **extensible**, **web-based**, and foster **reproducibility**. We have developed a case-study application called intRo which we will use to illustrate our method of developing a system meeting these criteria. The paper is structured as follows: Section \ref{case-study-intro} describes the motivation behind intRo as an extensible system. section \ref{what-is-intro} introduces the application, its features and its usability, and provides motivations for the design decisions underlying it. Section \ref{understanding-intro} provides technical details on how we build intRo, and how it can be used to develop other modular systems. Section \ref{deploying-intro-instances} describes the process of deploying a classroom instance of `intRo` tailored to the individual instructor. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of both intRo, and modular systems in general.

@@

Multiple software packages have recently been written in an attempt to spur interest in  R programming and statistics. DataCamp's [@datacamp] courses are user-friendly ways to learning basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics.

These limitations led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of access and an aim to foster student interest in coding. Attempting to entirely hide the programming aspect from students, even in introductory classes, is a lost opportunity to get students interested in statistical computing. It is also a lost opportunity reaching students who learn differently or have a computational background. Another advantage is its modular structure, which allows course instructors to tailor the application towards the needs of a particular class, rather than accept a piece of software as is. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text.

# What is `intRo`?

`intRo` is a web-based tool to accompany an introductory statistics class. It is meant to assist in the learning of statistics rather than as a stand-alone delivery of statistics education, with the intention of being used in conjunction with a guided class. An accompanying R package, titled `intRo` and available on GitHub, assists in the downloading, running, and deploying of `intRo` instances.

@@

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the student's R script. By default, the output will include the R code, but if the student elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results.

This reproducibility framework has three key advantages. First, this eases a student who may be intimidated by programming into the idea that interacting with a user interface is really just a frontend for code. Seeing the correspondence between graphical clicks and printed code will hopefully also lessen the fear of coding that many students may have. Second, an analysis created by `intRo` can be reproduced in an R session to easily assess and extend the results. Finally, "printing" the results of an `intRo` analysis amounts to nothing more than executing the R code on the server, adding another layer of reproducibility. These concepts are important because they encourage best practices with regards to disclosure of analysis methods in research [@baggerly2011reproducible; @xie2015].

@@

# Understanding `intRo`

To fully understand `intRo`, it is important to expand upon two underlying concepts. The first is the idea of modularity in the context of a Shiny application. The second is the idea of reactive programming in the context of a web application.

## Modules

@@

## Reactive design choices

@@

# Deploying `intRo` Instances

@@

>>>>

With the advantages of R clear, an approach to building statistical software and statistical learning tools would be to attempt to generate interest in programming, which could help naturally ease the transition into the use of R. Multiple software packages have recently been written in an attempt to spur this interest in R programming and statistics. DataCamp's [@datacamp] courses are user-friendly ways to learning basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics.

Modern web technologies have enabled a new generation of software packages that reside solely on the web, which eliminates the issue of local installation and helps abstract away some of the more challenging programming aspects of working directly with R. Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. Several recent software packages have built up Shiny to provide a web-based system based on R. One such package is iNZight Lite [@inzight] which attempts to expose students to data analysis without requiring programming knowledge. Like most web-based systems, this does not include reproducible R code, which limits its useful in a scientific or academic setting. Another package is called Radiant [@radiant]

Though challenging in a Graphical User Interface, a reproducibility framework has three key advantages. First, it eases a student who may be intimidated by programming into the idea that interacting with a user interface is really just a frontend for code. Seeing the correspondence between graphical clicks and printed code should help lessen the fear of coding that many students may have. Second, an analysis created by a reproducible software system can be brought in an R session to easily assess and extend the results. Finally, with the help of knitr and RMarkdown, "printing" the results of a reproducible software system analysis amounts to nothing more than executing the R code on the server, adding another layer of reproducibility. These concepts are important because they encourage best practices with regards to disclosure of analysis methods in research [@baggerly2011reproducible; @xie2015].

Based on this, we believe a modern software system should be **modular**, **extensible**, **web-based**, and foster **reproducibility**. We have developed a case-study application called intRo which we will use to illustrate our method of developing a system meeting these criteria. The paper is structured as follows:. Section \ref{case-study-intro} introduces the application, its features and its usability, and provides motivations for why it was built. Section \ref{intro-design-decisions} provides technical details on how we built intRo, by walking through the underlying modularity, reproducibility, and reactive framework, as well as how it can be used to develop other software systems with these properties. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of both intRo, and modular systems in general.

@@

These challenges led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of access and an aim to foster student interest in coding. Attempting to entirely hide the programming aspect from students, even in introductory classes, is a lost opportunity to get students interested in statistical computing. It is also a lost opportunity reaching students who learn differently or have a computational background. Another advantage is its modular structure, which allows course instructors to tailor the application towards the needs of a particular class, rather than accept a piece of software as is. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. An accompanying R package, titled `intRo` and available on GitHub, assists in the downloading, running, and deploying of `intRo` instances (See section \ref{deploying-intro-instances}).

@@

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable, reproducible R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the student's R script. By default, the output will include the R code, but if the student elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results.

@@

# intRo Design Decisions

In this section, we detail the design choices surrounding `intRo`'s extensibility. We have designed it in such a way that these ideas can be used in other Shiny-based software projects.

## Designing for Modularity

@@

## Designing for Reproducibility

## Designing for Reactivity

@@

@@

## Deploying `intRo` Instances
